Notes on Design and Implementation of Anglican

### 2015-07-29 paper/paper.tex
stochastic memoization

\subsubsection{Stochastic memoization}

>>>>

There are two proper probabilistic forms turning Anglican into a
probabilistic programming language --- \texttt{sample} and
\texttt{observe}. Their purpose is to interrupt deterministic
computation and transfer control to the inference algorithm.
Practically, this is achieved through returning
\textit{checkpoints} --- Clojure records of the corresponding
types (\texttt{anglican.trap.sample} or
\texttt{anglican.trap.observe}). The records contain fields
specific to each form, as well as the continuation; calling the
continuation resumes the computation. Checkpoints expose the
program state to the inference algorithm, and the updated state
is re-injected into the computation when the continuation is
called.

In addition to checkpoints, there are a few other special forms
--- \texttt{predict}, \texttt{store}, \texttt{retrieve},
\texttt{mem} --- which modify program state. These forms are
translated into expressions involving calls of functions from the
\texttt{anglican.state} namespace. The \texttt{mem} form, which
implements stochastic memoization, deserves a more detailed
explanation.

\textit{Memoization} is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the return value. There are no mutable data structures in a
probabilistic program, and \texttt{mem}'s memory is stored in a
nested dictionary in the program state (function
\texttt{mem-cps}).  Each memoized function receives a unique
automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation that
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state.

Memoized results are not shared among multiple runs of a
probabilistic program, which is intended. Otherwise, it would be
impossible to memoize functions with random results.
